---
title: "Assignment 3 Report"
format: html
editor: visual
---

## Results {#sec-report-results}

```{r}
#| label: fig-pa_group_GPA
#| fig-cap: GPA distribution by physical activity group
#| fig-pos: "H"
#| fig-align: "center"

ggplot(data2, aes(x = pa_group, y = GPA)) +
  geom_jitter(width = 0.2, 
              alpha = 0.3, 
              color = "steelblue") +
  stat_summary(fun = mean, 
               geom = "crossbar", 
               width = 0.4, 
               color = "red") +
  labs(x = "Physical Activity Duration Group",
       y = "GPA") +
  theme_minimal()  +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
```

```{r}
#| label: fig-pa_group_stress
#| fig-cap: Stress level distribution by physical activity group
#| fig-pos: "H"
#| fig-align: "center"

stress_distribution <- data2 %>%
  group_by(pa_group, Stress_Level) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = Stress_Level, values_from = count, values_fill = 0)

stress_prop <- stress_distribution %>%
  pivot_longer(cols = -pa_group, names_to = "Stress_Level", values_to = "count") %>%
  group_by(pa_group) %>%
  mutate(prop = count / sum(count)) %>%
  ungroup() %>%
  mutate(Stress_Level = factor(Stress_Level, levels = c("High", "Moderate", "Low")))

ggplot(stress_prop, aes(x = pa_group, y = prop, fill = Stress_Level)) +
  geom_bar(stat = "identity") +
  labs(x = "Physical Activity Group",
       y = "Proportion",
       fill = "Stress Level") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()  +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
```

@fig-pa_group_GPA presents the distribution of GPA across four physical activity groups. While individual GPAs vary substantially within each group, the mean GPA (marked in red) shows a slight downward trend as physical activity duration increases. Students exercising under 2 hours daily achieved the highest average GPA, while those in the more-than-6-hour group reported the lowest, suggesting a potential inverse relationship between physical activity and academic performance.

@fig-pa_group_stress displays the relative proportions of stress levels for each group. A clear shift is observed: students with low physical activity levels tend to report higher stress, whereas those engaging in 4-5 hours or more than 6 hours of physical activity are more likely to report moderate or low stress. The more-than-6-hour group has the highest proportion of low-stress individuals and the lowest proportion for high stress.

Together, these figures suggest that while higher physical activity may contribute to stress reduction, it might also correlate with a slight drop in GPA. These opposing trends point to a potential trade-off between academic focus and well-being.